---
title: "Elements of Probability"
output: html_notebook
---


# the coin toss game
## question 1
the probability of drawn scores for 2 throws
```{r}
dbinom(x=1, size=2, prob=0.5)
```

## Question 2
find the probability that after 10 throws our scores happen to be equal
- size = the number of throws = 10
- x = 5, the score (which must be equal, it is size divided by 2)
- prob = 0.5 because there is a 50/50 chance in a fair coin toss
```{r}
dbinom(x=5, size=10, prob=0.5)
```

- use the ```choose(n,k)``` function to calculate the binomial coefficient
- use the ```factorial()``` function to calculate factorials

## Question 3
(ref wk1 lecture 1 23:00)

What is the probability that after 11 throws my score is 3 points higher than yours?
- size = 11
- x = 4, because in order to meet the condition my score must be 7 and your score 4
```{r}
dbinom(x=4, size=11, prob=0.5)
```

# The tennis game

```{r}

```

# The birthday problem
## Question 1:
suppose there are n people gathered in a room. FInd the probability that at least 2 people have the same birthday when n people are gathered in a room

- nb: look at the compliment of the event rather thant the event itself
- need to give an algebraic answer

```{r}
k <- 100  # number of people in room
p <- numeric(k)  # create numeric vector to store probabilities
for (i in 1:k)      {
  q <- 1 - (0:(i - 1))/365  # 1 - prob(no matches)
  p[i] <- 1 - prod(q)  }
plot(p, main="Probability at least 2 people have same Birthday", xlab ="Number of People", ylab = "Probability", col="blue")
```

## question 2
what is the probability that all birthdays are distinct?

```{r}

```

```{r}
n <- 22
1 - prod(365: (365-n+1))/365^n
```


## question 3
what is the probability of 2 people having the same birthday if there are 23 people in the room?
```{r}
n <- 23
1 - prod(365: (365-n+1))/365^n
```

## question 4:
find the probability that at least one person has the same birthday as you, and find the value of n so that this probability approaches 1/2
- complimentary event (no one has my birthday)
```{r}
a <- 364
b <- 365  # the number of days in the year
p <- 0.5  # the probability
n <- log(p)/log(a/b)
n  # the number of people you need to have a 50% chance of
```

# conditional probabilities: aces in a deck

## Question 1
the probability of being dealt 4 aces
- n=52 because there are 52 cards in a deck
- k=4 because there are 4 aces in the deck
```{r}
a <- choose(52, 4)
a  # the number of distinct groups of 4 cards
1/a  # the probability of being dealt 4 aces
```

# question 2
four cards will again be dealt from a well-shuffled deck, and we now calculate the probability of 4 aces conditional on i aces having already been dealt
```{r}
1/choose(52-1, 4-1)  # i=1
1/choose(52-2, 4-2)  # i=2
1/choose(52-3, 4-3)  # i=3
```


# coded telegraph message
When a coded telegraph message is sent, there are sometimes errors in transmission. In particular, Morse code uses ”dots” and ”dashes”, which are known to occur in the proportion of 3:4.
- P(dot sent) = 3/7
- P(dash sent) = 4/7

```{r}
sent <- 3/7
received <- 7/8
notsent <- 1-sent
notreceived <- 1-received
denom <- sent * received + notsent * notreceived
(sent * received)/denom  # the probability of correctly receiving a dot

```

# colour blind
Suppose that 5% of men and 0.25% of women are colour-blind. A person is chosen at random and that person is colour-blind. Find the probability that the person is male. Assume males and females to be in equal numbers.

Let M and F denote the event that the randomly selected person is male or female, respectively. Let C denotes the event that the randomly selected person is colour-blind. From the question, we know the following probabilities:
- P(C|M) = 0.05
- P(C|F) = 0.0025
- P(M) = P(F) = 0.5

```{r}
PCM <- 0.05  # probability that the person is male and colour blind
PCF <- 0.0025  # probability that the person is femal and colour blind
PMF <- 0.5  # probability for each gender
PC <- (PCM * PMF) + (PCF * PMF)  # the probability that a person is colour blind
(PCM*PMF)/PC  # the probability that the person is male
```

# Telegraph signal
Consider telegraph signals ”dot” and ”dash” sent in the proportion 3 : 4, where erratic transmissions cause a dot to become a dash with probability 1/4 and a dash to become a dot with probability 1/3. If a dash is received, find the probability that a dash has been sent? (2 dp.)



```{r}

a <- 3/7  # P(dot sent)
b <- 4/7  # P(dash sent)
c <- 1/4  # P(dash rec | dot sent)
d <- 1/3  # P(dot rec|dash sent)
e <- 3/4  # P(dot rec|dot sent) implied
f <- 2/3  # P(dash rec|dash sent) implied
(f*b) / ((f*b) + (c*a))  # the probability that a dash was sent when a dash is received
```

# question
Assuming independence between signals, if the message dot-dot was received, find the probability distribution of the four possible messages that could have been sent.

1) conditional probability for dot-dot
```{r}
dd <- (e*a) / ((d*b)+(e*a))  # probability dot-dot sent and dot-dot received
dd
dd^2
```

2) conditional probability for dash-dot
```{r}

```


3) conditional probability for dot-dash

```{r}

```

4) conditional probability for dash-dash
```{r}

```


# Rodent Problem
Two litters of a particular rodent species have been born, one with two brown-haired and one grey-haired (litter 1), and the other with three brown-haired and two grey-haired (litter 2). We select a litter at random and then select an offspring at random from the selected litter. Find the probability that the animal chosen is brown-haired. (2dp.)

```{r}
l <- 1/2  # probability of litter 1 or 2
b <- 2/3  # probability of brown hair in litter 1
c <- 3/5  # probability of brown hair in litter 2
x <- (b*l) + (c*l)  # the probability that the randomly chosen animal is brown haired
x
```

Given that a brown-haired offspring was selected, find the probability that the sampling was from litter 1. (2dp.)
```{r}
(b*l)/x  # the probability that the animal was from litter 1
```

