---
title: "Doing statistics with R"
output: html_notebook
---
# Some handy functions
There are some general functions that are handy when doing statistics.

## sums
You can use sum to add the elements of one or more numerical vectors:

```{r}
sum(1:10)
sum(c(1.2, 2.3, 3.4))
sum(1:2, 3:5) # Two vectors
```

If there are missing values then the sum is unknown, so the result is NA:
```{r}
sum(1:5, NA)
```

You can tell the sum function to ignore missing values by adding the argument na.rm=TRUE.
```{r}
sum(1:5, NA, na.rm=TRUE)
```

## Differences
You can get the lagged differences of the elements in a numeric vector using diff. For each element, starting from the second, the difference between it and the previous element is calculated. Notice that the resulting vector has one fewer elements.
```{r}
x <- c(2, 4, 5, 4, 6, 8) # Six elements
diff(x) # Five elements
```

# cumulatives
```{r}
cumsum(c(2, 3, 1, 4, 2, 3, 4, 5, 6, 0, 4))
```

Similarly, you can get the cumulative product, cumulative maximum, and cumulative minimum:
```{r}
x <- c(2, 3, 1, 4, 2, 3, 4, 5, 6, 0, 4)
cumprod(x)
cummax(x)
cummin(x)
```

# Factorials
If n is an integer with n ≥\ge≥ 1 then the factorial of n is defined to be $1 \times 2 \times ... \times n$. R has a function factorial that calculates factorials:

```{r}
factorial(10)
```

Let's check this by calculating the factorial explicitly:
```{r}
# One way
1 * 2 * 3 * 4 * 5 * 6 * 7 * 8 * 9 * 10

# Another way
cumprod(1:10)[10] # Get the last cumulative product
```

The factorial function is vectorised, which means that you can apply it to a vector of integers and get the factorial of each integer in the vector:
```{r}
factorial(1:5)
```

# Combinations
Suppose you choose a sample of n objects from a population of N objects, without replacement. This sample is called a combination. How many such combinations are there? Answer:

$\{N \choose n} = \dfrac{N!}{n!(N-n)!}$

R has a function choose that calculates this number:

```{r}
choose(10, 3)
```

Let's check this, by calculating the number explicitly:
```{r}
factorial(10)/(factorial(3)*factorial(10 - 3))
```

If you'd like to actually generate the combinations of a given length from a vector then you can use combn:
```{r}
x <- c(1, 2, 3, 4, 5)
combn(x, 3)
```
# Random samples
The sample function gives you a way to generate random samples of data.
```{r}
# A random shuffle of the first 30 positive integers
sample(30)

# A random sample of 10 taken from the first 30 positive integers, without replacement
sample(30, size=10)

# A random sample of 10 taken from the first 30 positive integers, with replacement
sample(30, size=10, replace=TRUE)
```

# Summarising Data
Suppose you have some numerical values in a vector. R has a variety of functions to help you summarise the values.

# Mean and Median
You can use the functions mean and median to find the mean and median of the values:
```{r}
x <- sample(100, size=20, replace=TRUE)
x
mean(x)
median(x)
```

If there are any NAs in the vector then the results will be NA:
```{r}
# With NAs
mean(c(1:10, NA))
median(c(1:10, NA))
```

You can tell R to ignore the NAs by adding na.rm=TRUE:
```{r}
# Ignore the NAs
mean(c(1:10, NA), na.rm=TRUE)
median(c(1:10, NA), na.rm=TRUE)
```

If you'd like to exclude outliers from the calculations then you can add a trim argument. If you add trim=0.1 then R will sort the elements and ignore the first 10% and the last 10% of them. So, if the vector has 20 elements then it will ignore the first 2 and the last 2.
```{r}
x <- sample(100, size=20, replace=TRUE)
x
mean(x, trim=0.1)
median(x, trim=0.1)
```

In many cases the numerical vector in which you are interested is a column of a data frame:
```{r}
df <- read.csv('../data/wine.csv')
# Without ignoring NAs
mean(df$X1.04)
median(df$X1.04)

# Ignoring NAs
mean(df$X1.04, na.rm=TRUE)
median(df$X1.04, na.rm=TRUE)
```

# Range, variance, and standard deviation
You can use min and max to find the minimum and maximum values of a numerical vector:




```{r}
mycars <- within(mtcars, { cyl <- ordered(cyl) })
mycars
```


