---
title: "The R language"
output: html_notebook
---
# What is R?
R is a general purpose programming language, like many other general purpose programming languages such as Python, C,
PHP, Javascript, and so on. It is particularly well-suited to data manipulation and statistical analysis, because it was
originally designed by statisticians with these things in mind. It has thus become very popular among data analysts
and statisticians.

R was created by Ross Ihaka and Robert Gentleman at the University of Auckland. It is an implementation of the S
programming language, created by John Chambers at Bell Labs in 1976. R was first conceived in 1992, with an initial
version released in 1995 and a stable beta version released in 2000. It is now developed by the R Development Core Team.

R is named partly after the first names of the two R creators, and partly as a play on the name "S". There are some
important differences between R and S, but much of the code written for S runs unaltered in R.

# Why use R?
R’s syntax makes it easy to create complex statistical models with just a few lines of code. Also, it has a large
ecosystem of packages and other resources. Most notably for data science is the tidyverse, a group of packages built
specifically to make working with data in R quicker and easier. And, there is a strong community of R users, so it's
easy to find answers to question and to get guidance.

Many companies, banks, financial institutions, universities and research labs have data scientists using R, doing such
things as behavioural analysis, measurement of ad effectiveness, economic forecasting, and data analysis and
visualisation.

Many companies, banks, financial institutions, universities and research labs have data scientists using R, doing such
things as behavioural analysis, measurement of ad effectiveness, economic forecasting, and data analysis and
visualisation.

Even if you won't be using R yourself, learning the basics will allow you to understand and follow R code that you might
come across.


```{r}
2 + 2
log(20)
```

# Your working directory
You might find yourself needing to open files and save files. It can be helpful to set a default directory. This is
called your working directory. You can get and set your working directory by using the following commands:

```{r}
# # Get your working directory
# getwd()
#
# # Set your working directory
# setwd('/course/data/')
```

# Commands
When you use R you issue commands, which R interprets and executes. Here is a command - it asks R to calculate the sum of 2 and 2:


```{r}
2 + 2
```

You can combine many commands together into a script (aka program):

```{r}
2 + 2
16 * 7
(3*12) - (7*2)
```

# Functions
You can command R to execute a function. In the following command we ask R to calculate the log of 2

```{r}
log(2)
```
# Vectors
It might look as though you are working with numbers, but you are actually working with vectors of numbers, where a vector is a sequence of elements. The number of elements is its length. In the command 2 + 2, you are asking R to add a vector of length one whose element is the number 2, to a vector of length one whose element is the number 2. The result is also a vector - a vector of length one whose element is the number 4. This makes R a bit different from other programming languages. In other languages you work directly with numbers and characters etc. Not in R.

You can create longer vectors using the c function (short for "combine").

```{r}
c(2, 3, 4, 5)
```

When you add, subtract, multiply, or divide vectors, R performs the calculations element-wise:

```{r}
c(2, 3, 4, 5) + c(1, 2, 8, 10)
c(2, 3, 4, 5) * c(1, 2, 8, 10)
```

Similarly, when you apply a function such as log to a vector, R applies the function element-wise:

```{r}
log(c(2, 3, 4, 5))
```

# Variables
You can assign the result of a calculation or a function call to a variable, using the assignment command:

```{r}
x <- 2 + 2
```

Notice that assignment in R is done using a backwards arrow <-. In many other languages it is done using =.

Also notice that R doesn't print the result of the addition in this case. That's because it gets assigned to the variable. To see the result, you can do either of the following two things:

```{r}
# Print x after it gets its value
x <- 2 + 2
x

# Enclose the assignment command in parentheses
(x <- 2 + 2)
```

# Comments
You will have noticed the addition of comments above. You can add comments using #. Anything that comes after # on a
line is treated as a comment by R and ignored.

```{r}

```

# Scalars
As you've seen, in R you don't work directly with numbers, strings, etc. Rather, you work with vectors of them. Nevertheless, there are numbers, strings, etc. as elements of those vectors. It's useful to have a name for them - they are called scalars.

## Types of scalars
R distinguishes four types of scalars. Here they are, with some examples of how you write them in R:

- **Integer**: 1L, 24L, -6L, 12e6L. You have to add "L" at the end, to get R to treat a number as an integer. If you
don't add "L" then R will take it to be a double.
- **double**: 3.14, -67.25, 34e5. You can use scientific notation.
- **Character**: 'a', "Hello". You can use either single or double quotes, but they must match.
- **Logical**: TRUE, FALSE. You can abbreviate TRUE to T and FALSE to F, but it is not recommended.

There are also complex and raw, but these are less often used and we won't discuss these two types.

Integer and double scalars are both known as numeric scalars.

## Special scalars
There are four special scalars:
- Inf , -Inf (positive and negative Infinity)
- NaN (Not a Number)
- NA (Not Available)

Inf and -Inf are doubles. Inf is handy when you want a value that is guaranteed to be bigger than any number; -Inf is
handy when you want a value that is guaranteed to be smaller than any value. You can use them in comparisons and
calculations:

```{r}
Inf > 3745621312
-Inf < -45672346123
23/Inf
```

NA is a value that is used to represent the absence of a value (which makes it a bit paradoxical!). There is actually an NA of each scalar type - integer NA, double NA, character NA, and logical NA, for reasons that will be explained in a later slide.

# Objects

When you use R you work with objects. In the "Quick start" slide you met the two types of objects that you'll often work with:
- Vectors
- Functions

You might work with many different vectors - these are all different objects of the same type (vector). Similarly, you might work with many different functions - these are all different objects of the same type (function).

There is also a third type of object:
- NULL

There is actually only one object of the type NULL, and it's called NULL. The NULL object is used to represent the absence of an object (which makes it, like NA, a bit paradoxical!). Note that NULL is not the same thing as NA. NULL is an object used to represent the absence of an object; NA is a scalar used to represent the absence of a scalar.
Testing objects

You can test an object to see whether it is a vector, a function, or NULL, using the functions is.vector, is.function, and is.null:

```{r}
is.vector(c(1, 2, 3, 4, 5))
is.vector(log)
is.vector(NULL)

is.function(c(1, 2, 3, 4, 5))
is.function(log)
is.function(NULL)

is.null(c(1, 2, 3, 4, 5))
is.null(log)
is.null(NULL)
```

You need to be a little careful when using is.vector: for any vector that has an attribute other than names it returns FALSE. (This will be explained.)


```{r}

```

# Attributes
An important fact about R objects is that they can have attributes,  which you can think of as pieces of metadata about the object. R makes extensive use of attributes.

You can give an object any attribute you like, but R uses certain attributes for specific purposes, such as deciding how to display the object when it is printed, or how to apply functions to it.

Any object can have attributes, apart from NULL - it can't have any attributes.

## Setting and getting attributes
To give an object an attribute, and to later retrieve it, you can use the general purpose function attr:
```{r}
x <- c('Sydney', 'Brisbane', 'Melbourne')

# Set the value of an attribute
# Its name is 'tag' and its value is 'My favourite cities'
attr(x, 'tag') <- 'My favourite cities'

# Get the value of the attribute
attr(x, 'tag')
```

You can see what attributes an object has using the function attributes (if the object has no attributes the function returns NULL):

```{r}
x <- c('Sydney', 'Brisbane', 'Melbourne')
attributes(x)
attr(x, 'tag') <- 'My favourite cities'
attributes(x)
```

In the examples above we assign attributes to vectors, but you can assign them to functions too:
```{r}
attr(log, 'owner') <- 'me'
attributes(log)
```

# Special attributes
Some attributes have a special use in R. These include:
- **names**. The names of the elements of a vector.
- **dim**. The dimensions of a vector.
- **class**. The class of an object.
You will learn about these, and others, as we go.

Each special attribute has an **accessor function**, with the same name as the attribute. So, the special attribute ```names``` has an accessor function called ```names```, the special attribute dim has an accessor function called ```dim```, and so on. It is best to use these accessor functions when setting and getting attributes, rather than the general purpose ```attr```. You'll see how to do this in the next example.

## Names
By default, the elements of a vector do not have names. But you can give them names, by giving the vector a names attribute. Here's an example:
```{r}
x <- c(1, 2, 3)
attr(x, 'names') <- c('a', 'b', 'c')
attr(x, 'names')
```
Here's how you do it using the special accessor function ```names```:

```{r}
x <- c(1, 2, 3)
names(x) <- c('a', 'b', 'c')
names(x)
```

You can also specify a name for each vector when you use the ```c``` function - R automatically turns them into a ```names``` attribute. Notice that you don't have to use quotes around the names in this case:

```{r}
x <- c(a=1, b=2, c=3)
names(x)
```

Names don’t have to be unique. However, as you will see later, they can be a helpful way of referring to an element of a vector, so they are most useful when they are unique.

Not all elements of a vector need to have a name. If some names are missing when you create the vector, the names will be set to an empty string for those elements.

```{r}
x <- c(a=1, b=2, 3)
x
names(x)
```

If you modify the vector by setting some variable names, but not all, ```names``` will return ```NA``` for the missing ones.

```{r}

```

You can remove the names from a vector by setting the names attribute to NULL, or by using the function ```unname```:

```{r}
x <- c(a=1, b=2, c=3)
x
names(x) <- NULL
x

x <- c(a=1, b=2, c=3)
x
x <- unname(x)
x
```
# Vectors

Testing vectors: You can test a vector to see whether it is one of these types, using the following functions:

```{r}
is.atomic
is.list
is.data.frame
is.array
is.matrix
```

You need to be a little careful using ```is.atomic```, because it returns ```TRUE``` for the object ```NULL```, even though ```NULL``` is not an atomic vector (it's not even a vector):

```{r}
is.atomic(NULL)
```

Many people treat the NULL object as if it were an atomic vector of length zero, anyway, so usually this doesn't cause problems. If the difference matters to you then you can use the is.null function to distinguish NULL from atomic vectors.:

```{r}
is.null(NULL) # TRUE
is.null(c(1, 2, 3)) # FALSE
```

# Some vector functions
There are many functions for working with vectors. Many of them apply only to certain types of vectors, but some of them, including the following, apply to all vectors.

You can get the length of any vector using the ```length``` function:

```{r}
x <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
length(x)
```

You can see the first few elements of any vector using the head function, and the last few elements using the tail function. By default, it shows 6. If you'd like a different number, you can specify it as an argument.

```{r}
x <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
head(x)
head(x, 3)
tail(x)
tail(x, 3)
```

You can reverse the elements of any vector using the rev function:

```{r}
x <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
rev(x)
```
# Functions
Functions play a very important role in R, so it's good to have a solid understanding of them.
Calling functions

Every function has a name, such as ```c```, ```log```, or ```help```.  You can **call** a function (or **invoke** it) by adding parentheses (round brackets) after its name, inside which you can specify values for any **arguments** the function has. Arguments are separated by commas.

```{r}
log(2, 10)
```
Some argument values are required, some are optional. Some argument values can be named, some can't. Some argument values must be given in a certain position, some can move around. Part of understanding a function is knowing these things about its arguments.

You can learn about a function, such as ```log```, using either ```help(log)``` or ```?log```:
```{r}
?log
```
The information given is often not that easy to understand, but you should be able to find what you need. Under the section "Usage" it says the following:
```log(x, base = exp(1))```

Alternatively, you can use the args function, which just gives information about the arguments of the function:

```{r}
args(log)
```
The information we get tells us that the log function has two arguments, named "x" and "base".

You can supply values for these arguments with or without naming them:
```{r}
log(2, 10) # Without naming them
log(x=2, base=10) # Naming them
```
If you don't name them then you must provide the values in the correct order ```x``` first,  base second - the function will interpret the first value as a value for ```x```, and the second value as a value for ```base```.
```{r}
log(2, 10) # x=2, base=10
log(10, 2) # x=10, base=2
```
If you name them then you can put them in any order you like - it makes no difference:
```{r}
log(x=2, base=10)
log(base=10, x=2) # Same
```
So, although naming the values requires you to type more, it makes it easier for you to give the right values to the right arguments.

The ```x``` argument has no default value, which means that it is required - you must supply a value for this argument. The ```base``` argument has a default value of ```exp(1)```, (i.e. e, or 2.718...) which means that it is optional - you don't have to supply a value for this argument (the default value will be used if you don't), but you can if you want (in which case the value you supply will be used, rather than the default).
```{r}
log(2) # No value for base supplied - the default value of exp(1) is used
log(2, 10) # A value of 10 is supplied for base
```

# Return values
In R, when a function is called it always returns a value, at the very least the object ```NULL```. You can assign that value to a variable, in which case the value is not printed:

```{r}
x <- log(2) # Returned value is assigned to x and not printed
```
If you don't assign it to a variable then R prints it:

```{r}
log(2) # Returned value is not assigned to a variable so is printed
```

If you'd like to do both - assign the returned value to a variable and also print it - then you can put the assignment command inside parentheses:
```{r}
(x <- log(2)) # Returned value is both assigned and printed
```

# Functions are objects
Functions are objects, just like vectors are, and you can work with them just like you can work with vectors.
You can get R to print a function:
```{r}
log
```
You can assign a function to a variable, in which case the variable becomes another name for the function:
```{r}
x <- log
x(2, base=10)
```
And functions can be supplied as arguments to functions:
```{r}
help(log)
```
# Every command is a function call
A curious thing about R is that every command is actually a function call.
```{r}
# This command
2 + 3
# Is really this function call
`+`(2, 3)

# This command
x <- 10
# Is really this function call
`<-`(x, 10)
```
Why the backticks around the function names? Because these names are not legal R names - they don't conform to R's naming rules (these are explained in the next slide). You can still use them, but you have to surround them by backticks (actually, quote marks will do, but it's customary to use backticks).

If you'd like, you could work with R using nothing but the "function call" style of commands. But you'll probably find it easier not to. You'll probably find it more intuitive to use the command ```2 + 3``` rather than the command `+`(2, 3) (which is why the former is sometimes said to be **syntactic sugar** for the latter).

# Generic functions
You might sometimes hear a function referred to as a **generic function**. This means that the function is sensitive to the class attributes of any object it is supplied, and might do different things with objects of different classes.


# Variables
You won't get very far before finding it useful to introduce variables and assign values to them.

## Assignment
The standard way to assign a value to a variable is to use <-:
```{r}
x <- c(1, 2, 3, 4, 5)
x
```
This is not the only way. You can also use ```->``` and  ```=```:

```{r}
c(1, 2, 3, 4, 5) -> x
x

x = c(1, 2, 3, 4, 5)
x
```
There is even a function assign that you can use:

```{r}
assign('x', c(1, 2, 3, 4, 5))
x
```
In fact, when you use ```<-```, ```->``` or ```=``` behind the scenes R just converts these into calls of the assign function.

Despite having these various techniques available, it is recommended that you always use <-.

## Multiple assignments
You can make multiple assignments in one command, as follows
```{r}
x <- y <- z <- c(1, 2, 3, 4, 5)
x
y
z
```
This works because the assign function, in addition to making the assignment, returns the value that is assigned. So z <- c(1, 2, 3, 4, 5) assigns c(1, 2, 3, 4, 5) to z and returns c(1, 2, 3, 4, 5). This c(1, 2, 3, 4, 5) then gets assigned to y, and c(1, 2, 3, 4, 5) is again returned. This c(1, 2, 3, 4, 5) then gets assigned to x, and c(1, 2, 3, 4, 5) is again returned (but not used). In effect, it's equivalent to the following:
```{r}
x <- (y <- (z <- c(1, 2, 3, 4, 5)))
x
y
z
```
# Naming rules
There are rules governing what expressions you can use as a variable:
- Variables can only include alphanumerical characters (a-z, A-Z, 0-9), the underscore (_), or the dot (.) (so no white space)
- Variables cannot start with a digit
So the following are all valid variables:

```{r}
# x
# first_name
# mobile.phone.1
# mobile.phone.2
```

But these are not:
```{r}
# 1st           (Can't start with a digit)
# width*length  (Can't include *)
# $AU           (Can't include $)
# first-name    (Can't include -)
# Last Name     (Can't include a space)
# total%        (Can't include %)
```
Variables are case-sensitive. So the following are different variables:

```{r}
# name
# Name
```
If for some reason you would like to use a variable that violates the rules then you can do so, as long as you enclose it in backticks (actually, single or double quotes will work too, but backticks are recommended).
```{r}
# `*3` <- 10
# `*3`
#
# `my var` <- 10
# `my var`
```

# The dot
If you've worked with Python or other languages and are new to R, then you might get confused by the dot (.). In Python, and various other languages, the dot has special significance - you use it to access an object's attributes. Not so in R - the dot is just another character. It often gets used in R to separate words in a variable (e.g. first.name) , in much the same way that you might use an underscore in Python (e.g. first_name). Keep this in mind!

# Choosing variables
In addition to respecting the rules above, you should choose variables that are easy to read, easy to understand, and easy to remember.

Use variables that remind you what they stand for.

# Packages
The functionality that comes automatically with R is called **base R**. There is quite a bit in base R, but you will often find yourself wanting to use functionality that is kept in some other R package, not in base R.

A **package** is a collection of R functions, possibly with some data, and often with some documentation. The location where the packages are stored is called a **library**.

To use a package it needs to be **installed** on your computer, and you need to **load** it.

# Checking for installed packages
To check what packages are installed on your computer, you can use:

```{r}
installed.packages()
```
Installing packages

If you need a package that is not installed on your computer then you need to install it. How you install a package depends on where it is located. For publicly available packages, this means what repository it belongs to. Most commonly this is the CRAN repository, in which case you just need the name of the package and use the function install.packages.

You might find it helpful to install devtools, using install.packages. After devtools is installed you will be able to use its various utility functions to install another packages. The utility functions are:
- install_bioc, for installing from Bioconductor
- install_bitbucket, for installing from Bitbucket
- install_cran for installing from CRAN
- install_git, for installing from a git repository repository
- install_github, for installing from GitHub
- install_local, for installing from a local file
- install_svn, for installing from an SVN repository
- install_url, for installing from a URL
- install_version, for installing from a specific version of a CRAN package
- For example, to install the babynames package from its Github repository, you can use:

```{r}
# devtools::install_github("hadley/babynames")
```

# Updating packages
You can check which of your installed packages needs updating using old.packages:

```{r}
# old.packages()
```
You can update all packages using update.packages:
```{r}
# update.packages()
```


```{r}
install.packages("dplyr")
```

# Removing packages
You can uninstall a package using remove.packages:

```{r}
remove.packages("dplyr")
```

# Checking for loaded packages
To see which packages are currently loaded you can use search:
```{r}
search()
```

# Loading packages

Once a package is installed you can use its functionality. If you just need to use one of its items occasionally you can call it using the notation packagename::itemname. Suppose you have the package babynames installed. Then you can open its births dataset as follows:

```{r}
babynames::births
```

For more extensive use you will probably want to load the package. Then you don't need to keep using the package:: prefix.
There are two ways to load a package:
- Use library. This generates an error if the package cannot be found.
- Use require. This returns ```TRUE``` if the package is found, otherwise it returns ```FALSE``` (with no error).

Using ```library```:

```{r}
library("dplyr")
```

Using ```require```:
```{r}
require("dplyr")
```
It is best to use library, because you want to get an error if the package cannot be found.

Neither library nor require actually execute any code. They simply make the package available to use. Individual functions in the package are not run unless they are actually called later. The function source, on the other hand, executes all of the code in the package when it is called.

# Unloading a package
To unload a given package, you can use the detach function:
```{r}
library("dplyr")
detach("package:dplyr", unload=TRUE)
```
# Help with a package
If you'd like help with a package you can use the following:

```{r}
help(package="dplyr")
# Or
packageDescription("dplyr")
```

To see what is available to you in a package you can use ls:

```{r}
library("dplyr")
ls("package:dplyr")
```

To get help with a particular function in a package you can, as usual, use help or ?:

```{r}
library("dplyr")
help("mutate")
```

# Vignettes
Most packages have vignettes, which showcase the functionality of the package. To see what vignettes are available you can use vignette:

```{r}
vignette()
```
To see what's available within a particular package you can specify it:

```{r}
vignette(package="dplyr")
```

To see a vignette (this probably won't work here in Ed, but it will in RStudio):
```{r}
vignette("grouping")
```
