---
title: "Elements of Probability"
output: html_notebook
---
# the coin toss game

## q1
find the probability that after 10 throws our scores are equal

- size = the number of throws = 10
- x = 5, the score (which must be equal, it is size divided by 2)
- prob = 0.5 because there is a 50/50 chance in a fair coin toss
```{r}
dbinom(x=5, size=10, prob=0.5)
```

## q2
find the probability that after 11 throws my score is three points higher than yours

- size = the number of throws = 11
- x = 7, if my score is 3 higher than yours, and the total is 11, then my score must be 7
- prob = 0.5 because there is a 50/50 chance in a fair coin toss
```{r}
dbinom(x=7, size=11, prob=0.5)
```

- use the ```choose(n,k)``` function to calculate the binomial coefficient
- ref tutorial 1 @ 15:16
```{r}
choose()
```

- use the ```factorial()``` function to calculate factorials
```{r}
factorial()
```
# The Tennis Game

## q1 suppose there are 8 players, find the probability that they meet in the first or second round (2dp)
8 players
3 rounds
probability 0.5
```{r}
dbinom(x=3, size=8, prob=0.5)
```
# The birthday problem
there is supposed to be a version for an undefined n which i can't get working

## q1
suppose that there are n people gathered in a room. find the probability that at least two people have the same birthday when n people are gathered in a room.
```{r}

```
## Question 1: ref slide 17
suppose there are n people gathered in a room. FInd the probability that at least 2 people have the same birthday when n people are gathered in a room
- nb: look at the compliment of the event rather thant the event itself
- need to give an algebraic answer
```{r}
k <- 100  # number of people in room
p <- numeric(k)  # create numeric vector to store probabilities
for (i in 1:k)      {
  q <- 1 - (0:(i - 1))/365  # 1 - prob(no matches)
  p[i] <- 1 - prod(q)  }
plot(p, main="Probability at least 2 people have same Birthday", xlab ="Number of People", ylab = "Probability", col="blue")
```
## q2
calculate the probability that at least two people have the same birthday for n = 22 (2dp)
```{r}
n <- 22
1 - prod(365: (365-n+1)) / 365^n
```
## q3
calculate the probability that at least two people have the same birthday for n=23
```{r}
n <- 23
1 - prod(365:(365-n+1)) / 365^n
```
# q 4
find the probability that at least one person has the same birthday as you and find the value of n so that this probability approaches 1/2
```{r}
a <- 364  # the number of outcomes not in the event. the compliment.
b <- 365  # the number of days in the year
p <- 0.5  # the probability
n <- log(p)/log(a/b)
n  # the number of people you need to have a 50% chance that at least one person has the same name as you
```

# cards example
## question 1
four cards are deal from the top of a well shuffled deck of 52 playing cards. FInd the probability that they are 4 aces.

```{r}
groups <- choose(52, 4)
groups
1 / groups
```
## Question 2
four cards will again be dealt from a well-shuffled deck, and we now calculate the probability of 4 aces conditional on i aces having already been dealt

ref slide 25
```{r}
1/choose(52-1, 4-1)  # i = 1
1/choose(52-2, 4-2)  # i = 2
1/choose(52-3, 4-3)  # i = 3
```
# Coded Telegraph Message
When a coded telegraph message is sent, there are sometimes errors in transmission. In particular, Morse code uses ”dots” and ”dashes”, which are known to occur in the proportion of 3:4.
- P(dot sent) = 3/7
- P(dash sent) = 4/7

```{r}
sent <- 3/7
received <- 7/8
notsent <- 1-sent
notreceived <- 1-received
denom <- sent * received + notsent * notreceived
denom  # probability dot received
(sent * received)/denom  # the probability of correctly receiving a dot

```
# Colour Blind
## q1
Suppose that 5% of men and 0.25% of women are colour-blind. A person is chosen at random and that person is colour-blind. Find the probability that the person is male. Assume males and females to be in equal numbers.

Let M and F denote the event that the randomly selected person is male or female, respectively. Let C denotes the event that the randomly selected person is colour-blind. From the question, we know the following probabilities:
- P(C|M) = 0.05
- P(C|F) = 0.0025
- P(M) = P(F) = 0.5
```{r}
PCM <- 0.05  # probability that the person is male and colour blind
PCF <- 0.0025  # probability that the person is femal and colour blind
PMF <- 0.5  # probability for each gender
PC <- (PCM * PMF) + (PCF * PMF)  # the probability that a person is colour blind
PC  # probability that the person is colour-blind
(PCM*PMF)/PC  # the probability that the person is male
```
# Telegraph signal
## q1
Consider telegraph signals ”dot” and ”dash” sent in the proportion 3 : 4, where erratic transmissions cause a dot to become a dash with probability 1/4 and a dash to become a dot with probability 1/3. If a dash is received, find the probability that a dash has been sent? (2 dp.)

```{r}
a <- 3/7  # P(dot sent)
b <- 4/7  # P(dash sent)
c <- 1/4  # P(dash rec | dot sent)
d <- 1/3  # P(dot rec|dash sent)
e <- 3/4  # P(dot rec|dot sent) implied
f <- 2/3  # P(dash rec|dash sent) implied
(f*b) / ((f*b) + (c*a))  # the probability that a dash was sent when a dash is received
```
## q2
Assuming independence between signals, if the message dot-dot was received, find the probability distribution of the four possible messages that could have been sent.

conditional probability for dot-dot
```{r}
dd <- (e*a) / ((d*b)+(e*a))  # probability dot-dot sent and dot-dot received
dd
dd^2  # because of independence
```
# Rodents
## q1
two litters of a particular rodent species have been born, one with two brown-haired  and one grey-haired (litter 1), and the other with three brown haired and two grey-haired (litter 2). We select a litter at random and then select an ofspring at random from the selected litter. Find the probability that the animal chosen is brown-haired. (2dp.)

*ref slide 35*
```{r}
litter <- 1/2  # probability of litter 1 or 2
p1 <- 2/3  # probability of brown hair in litter 1
p2 <- 3/5  # probability of brown hair in litter 2
pbh <- (p1*litter) + (p2*litter)  # the probability that the randomly chosen animal is brown haired
pbh  # probability that the randomly chosen animal is brown haired
```
## q2
given that a brown-haired offspring was selected, find the probability that the sampling was from litter 1 (2dp)
```{r}
litter <- 1/2  # probability of litter 1 or 2
p1 <- 2/3  # probability of brown hair in litter 1
(p1*litter)/pbh  # the probability that the animal was from litter 1
```
# distributions
# geometric distribution
```{r}
rgeom(5, 0.5)
```
# poisson distribution
## q1
If there are twelve cars crossing a bridge per minute on average, find the probability of having seventeen or more cars crossing the bridge in a particular minute.

Let X be the number of accidents at a busy Sydney intersection in a month. The average number of accidents per month is three. Calculate the probability that zero accidents occur in a particular month.

q = the number of incidents we want to test for
lambda = the average number of accidents per month

we use ppois because...
```{r}
ppois(0, lambda=3)
```
# store traffic
## q1
Suppose that on average 5 customers walk into a store every minute. Find the probability that 10 customers come into the store the next minute (3dp)

q = 10
lambda = 5
```{r}
dpois(10, lambda=5)
```
## q2
find the probability that at least five customers come into the store the next minute (3dp)
q = 5
lambda = 5
we use dpois because...
```{r}
dpois(5, lambda=5)
```
# The Bernoulli Distribution
Suppose that X is the outcome of flipping a coin where a head is given the outcome one and a tail is given the outcome 0. COmpute the rpobability that a head is flipped and also that a tail is flipped

```{r}

```
# the binomial distribution

## q1
suppose that X is Binomially distributed with n=5 trials and probability of success p = 0.8. Compute the probability that there are four successful trials, ie: P(X=4)

ref slide 48
```{r}
dbinom(4, size = 5, prob=0.8)
```
## q2
suppose there are 10 multiple choice questions in a maths quiz, where each question has four possible answers and only one of them is correct. Hint, the value of p first.

probability = 0.25
size = 10
x = 5

ref slide 49
```{r}
dbinom(x=5, size = 10, prob=0.25)
```

# Cumulative Distribution Function

## q1 if the CDF of a distribution is

# the probability density function theorem

## question


# Common Continuous Distributions


# Transformations


# Density Transformation Formula


# Density Transformation example
example from slide 73

```{r}
## Plot the density functions
pdf <- function(y) 14*y*(1-y^(1/3))
curve(pdf,0,1)
```

show that the density function integrates to one
```{r}
integral <- integrate(pdf,0,1)
integral
```


## Example 2
suppose that X has the density function $f_X (x) = 7e^{-7x}  $0 < x < \infty$
```{r}
# Plot the density function
pdf <- function(y) (7/4)*exp(-(7/4)*(y-3))
curve(pdf,3,10)
```
Show that the density function integrates to 1
```{r}
integrate(pdf,3,Inf)
```
## q2
For the following $Y=X^2$ where $X$ has the density:
$f_X (x) = 30x^2(1-x)^2$,   $0 < x < 1$
- find the pdf of Y
- plot the pdf of Y in R
- Show that the pdf integrates to one in R
```{r}
pdf <- function(y) 15*sqrt(y)*(1 - sqrt(y))^2
curve(pdf,0,1)
```
show that the density function integrates to one
```{r}
integrate(pdf,0,1)
```

# Probability Integral Transform
ref slide 80
```{r}
data <- -log(1-runif(10000))
hist(data, prob = TRUE, breaks = 100)
curve(exp(-x), add = TRUE, col = "purple", lwd=2)
```

# R's built-in distributions
R provides built in random number generators for well known distributions. For example, if you wish to generate observations from the Gamma(a,b) distribution, you can do the following:

```{r}
n = 10000
alpha = 0.8
beta = 5
sim.dat = rgamma(n, alpha, beta)
hist(sim.dat, prob = TRUE, breaks = 100)
curve(dgamma(x,alpha,beta),add=TRUE,col="orange", lwd = 2)
```

# expected Values
Suppose X has an exponential λ distribution, with density

$f_X(x) =\dfrac{1}{\lambda}e^{-x/\lambda}, $0 \leq x < \infty, \lambda > 0$.

Find the expected value of X.

```{r}
pdf <- function(x) 1/l * e^(-x/l)
E <- integrate(pdf, lower = 0, upper = Inf)
E
```
# q
Apple is planning to host a live webcast of the next wwdc conference. The server has a maximum capacity of 3 million people (measured in millions) that want to watch the live webcast, with pdfs

$f(x) = \dfrac{x^2}{9}$ for $0 < x < 3$

this one is problematic
```{r}
pdf <- function(x) (x^2)/9
mean(pdf)
```

## q
find the variance of the number of people who want to watch the webcast. Give your answer to 4 decimal places

```{r}

```

# q
show that if a pdf is symmetric about a point a, the $\alpha_3 = 0$. Hint: Show that $\mu_3 = 0$ for a general density function $f(x)$. note that $\mu = a$

```{r}

```

# q
calculate $\alpha_4$ for the following:

$f(x) = \dfrac{}{}, -1 < x < 1$

- first calculate the mean (the first momement)
- second calculate the variance (the second moment)
- the calculate the third and fourth central moments
- skewness is $\alpha_2$
- kurtosis is $\alpha_4$

```{r}

```


# Covariance and correlation
will be discussed in the next unit on multivariate distributions