---
title: "Multivariate Distributions"
output: html_notebook
---

# joint cumulative distribution function

# joint probability mass function

# Joint Density Function

## q1
a pdf is defined by:
$$f(x,y) = \begin{cases}
c(x+2y) & \text{ if } 0 < y < 1 \text{ and } 0 < x < 2 \\
0 & \text{ otherwise}$$

Find the value of C

define the function
```{r}
df <- function(x)
df
```

the density function must integrate to 1 over the range of x and y
```{r}
llimy <- 0; llimx <- 0
ulimy <- 1; ulimx <- 2

f <- function(x,y) (x+2*y)

integrate(function(y) {
    sapply(y, function(y) {
      integrate(function(x) f(x,y), llimx, ulimx)$value
    })
  }, llimy, ulimy)
```
therefore 4c = 1
hence c = 0.25


```{r}
c <- 1/4
c
```

## q2
find the joint cdf of X and Y
```{r}
llimy <- 0; llimx <- 0
ulimy <- 1; ulimx <- 2

f <- function(x,y) (x+2*y)

integrate(function(y) {
    sapply(y, function(y) {
      integrate(function(x) f(x,y), llimx, ulimx)$value
    })
  }, llimy, ulimy)
```
## q3
compute the probability #P(X \leq 1, Y \leq 0.25) to 2 decimal places

```{r}

```
## q4
Let X and Y be independent random variables with the same geometric distribution (same parameter p) for p=0.5, the pmf looks like the following:
```{r}

```
# q5 what is the joint pmf of (X,Y)? Hint, the joint pmf of independent random variables is given by
$P(X=x, Y=y) = P(X=x)P(Y=y)$


```{r}

```

# marginal distributions
## q1
consider the pdf:

Find the marginal distribution of X
```{r}

```
# Conditional Distribution


```{r}

```
# Moments and transformations


```{r}

```
# Density Transformation Formula


```{r}

```
## q1
let $X_1$ and $X_2$ have the following joint density function $f(x_1, x_2) = 1$ defined over the unit square $0 \leq x_1 \leq 1$. We will now consider the transformed variables $Y_1 = X_1 + X_2$ and $Y_2 = X_1 - X_2$. Find the joint pdf of $Y_1$ and $Y_2$

```{r}

```

#
given the pdf

find the pdf of the random variable $Z = 9/(X+1)^2$

- first calculate the marginal density
- apply the density transformation formula

```{r}
pdf <- function(x,y) x+2*y
integrate(pdf,0,2)
```
# q
Let $X_1$ and $X_2$ have joint pdf

$$

find the joint pdf of $Y_1 = X_1^2$ and $Y_2 = X_1X_2$

ref slide 125
```{r}

```

