---
title: "Random Variables and Univariate Distributions"
output: html_notebook
---

# Random Variables example

```{r}
library(ggplot2)

headstails <- data.frame(s=c("HHH","HHT","HTH","THH","TTH","THT","HTT","TTT"),
                         X=c(3,2,2,2,1,1,1,0))

ggplot(data = headstails, aes(x = X, y = ..density..)) +
  geom_histogram(bins=4,  fill="royalblue", col="grey") +
  scale_y_continuous(breaks = seq(0, 1, by = 0.05))
```

visualise the probability mass function for the parameter $p=0.6$
```{r}
library(ggplot2)

x <- 1:8
p <- 0.6
P <- (1-p)^(x-1)*p
dat <- data.frame(x,P)
ggplot(data=dat, aes(x=x, y=P)) +
  geom_bar(stat="identity", width=0.1,fill="orange2") +
  scale_x_continuous(breaks = seq(0, 10, by = 1))
```

on average there are 5 customers that walk into the store every minute

$\lambda = 5$

# Question 1
find the probability that 10 customers come into the store the next minute (3dp)

$x = 10$

**Hint**: try the function ```dpois(x, lambda)``` which calculates $P(X = x)$ when $X \sim \text{Poisson}(\lambda)$

```{r}
# Using exact probability
dpois(x=10, lambda=5)
```

# Question 2:
Find the probability that at least 5 customers come into the store in the next minute (3dp)

**Hint**: Try the function ```ppois(x, lambda)``` which calculates $P(X \leq x)$ when $X \sim \text{Poisson}(\lambda)$

```{r}
1 - ppois(q=4,lambda=5)
```

# Binomial Distribution

Suppose there are 10 multiple choice questions in a math quiz, where each question has four possible answer and only one of them is correct

**Hint**: find the value of p first

## Question 1:
find the probability of having exactly 5 correct answers (3dp)

**Hint**: try the function ```dbinom(x,n,p)```

because one out of four options is correct, p=0.25

```{r}
dbinom(x=5, size=10, prob=0.25)
```

# the cumulative distribution function (cdf)

```{r}
library(ggplot2)

headstails <- data.frame(s=c("HHH","HHT","HTH","THH","TTH","THT","HTT","TTT"),
                         X=c(3,2,2,2,1,1,1,0))
P <- ecdf(headstails$X)

P(0)
#[1] 0.125
P(1)
#[1] 0.5
P(2)
#[1] 0.875
P(3)
#[1] 1

ggplot(headstails , aes(x=X)) + stat_ecdf(geom = "step", colour='royalblue')

```

# cumulative distribution function example

```{r}
# sample size
n <- 1000

# random geometric data
x.geom <- rgeom(n=n, prob=.3)

# Cumulative Distribution Function
plot(ecdf(x.geom),
     xlab = "x",
     ylab = "P(X <= x)",
     col = "royalblue",
     cex = 0.3,
     lwd = 0.2,
     pch = 21,
     main = "Cumulative Distribution Function \n Geometric(n=1000, p=0.3)")
```


# student t distribution

```{r}
library(ggplot2)
x <- seq(-7, 7, by = 1/8)		# Range of x
y1 <- dt(x, df = 2)				# t distribution for 2 degrees of freedom
y2 <- dt(x, df = 4)				# t distribution for 4 degrees of freedom
y3 <- dt(x, df = 8)				# t distribution for 8 degrees of freedom
y4 <- dt(x, df = 16)			# t distribution for 16 degrees of freedom
y5 <- dt(x, df = 32)			# t distribution for 32 degrees of freedom
plot(x, y1,type="l",xlab="X",ylab="t distribution", ylim=range(seq(0,0.4,by=0.1)),col="blue")
lines(x, y2,type="l", col="red")
lines(x, y3,type="l", col="cyan")
lines(x, y4,type="l", col="green")
lines(x, y5,type="l", col="yellow")
legend("topright", legend=c("v=2","v=4", "v=8", "v=16", "v=32"),
       col=c("blue","red", "cyan", "green", "yellow"), lty=1)
```


# chi squared distribution

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)

data.frame(chisq = seq(0, 25, by=0.01)) %>%
  mutate(df_2 = dchisq(x = chisq, df = 2),
         df_4 = dchisq(x = chisq, df = 4),
         df_8 = dchisq(x = chisq, df = 8)) %>%
  gather(key = "df", value = "density", -chisq) %>%
  ggplot() +
  geom_line(aes(x = chisq, y = density, color = df)) +
  labs(title = "Chi-Square distribution for various Degrees of Freedom",
       x = "x",
       y = "pdf")
```
```{r}

```

# transformations and expectations example
```{r}
library(ggplot2)

base <- ggplot(data.frame(x = c(0,1)), aes(x))
myfunc <- function(x) -log(x)
base + stat_function(fun = myfunc, colour = "blue") +
  scale_x_continuous(breaks = seq(0, 1, by = 0.2)) +
  ylim(0,10)
```


```{r}
## Plot the density function
pdf = function(y) 14*y*(1-y^(1/3))
curve(pdf,0,1)
## Show that the density function integrates to one
integrate(pdf,0,1)
```

suppose that $X$ has a density function

$$f_X(x) = 7e^{-7x},  0<x<\infty$$

find the pdf of $Y = 4X + 3$ then using R plot the density function and show that the pdf integrates to one

```{r}
## Plot the density function
pdf <- function(y) (7/4)*exp(-(7/4)*(y-3))
curve(pdf,3,10)
## Show that the density function integrates to one
integrate(pdf,3,Inf)
```

# plotting probability density function question 2
for the following $Y=X^2$ where $X$ has the density:

$$f_X(x) = 30x^2(1-x)^2,   0<x<1$$

- find the pdf of $Y$
- plot the pdf of $Y$ in R
- show that the pdf integrates to one in R

```{r}
## Plot the density function
pdf <- function(y) 15*sqrt(y)*(1-sqrt(y))^2
curve(pdf,0,1)
## Show that the density function integrates to one
integrate(pdf,0,1)
```

# answer:
The density of $Y$ is

$$f_Y(y)=15(1- \sqrt{y})^2 \sqrt{y}$$
$$S_Y = [0,1]$$


# the exponential density

```{r}
data <- -log(1-runif(10000))
hist(data, prob = TRUE, breaks = 100)
curve(exp(-x), add = TRUE, col = "purple", lwd = 2)
```

# R's built in distributions
try experimenting with the random number generation using the R built-in distributions

```{r}
n <- 10000
alpha <- 0.8
beta <- 5
sim.dat <- rgamma(n, alpha, beta)
hist(sim.dat, prob = TRUE, breaks = 100)
curve(dgamma(x,alpha,beta),add=TRUE,col="orange", lwd = 2)
```

# moments

```{r}
# install.packages("moments")
library(moments)
data <- c(88, 95, 92, 97, 96, 97, 94, 86, 91, 95, 97, 88, 85, 76, 68)
```


```{r}
# calculate skewness
skewness(data)
```


```{r}
# calculate kurtosis
kurtosis(data)
```

# (from the lecture)
```{r}
choose(2,1)
```
```{r}
factorial(n)
```


```{r}
choose(52,4)
```

# telegraph example 1
```{r}

```


#
```{r}
sqrt(1/3)
```
