---
title: "Assessment 1 (10%)"
output: html_notebook
---
Due: 12pm midday Monday week 2

### Question 1:
suppose that three unbiased dice are rolled. Find the probability that at least two of the three dice have a face value
of 5 of more. Give your answer to 2 decimal places. we need to add the probability that x=2 and the probability that x=3

```{r}
# dbinom(x=2,size=3, prob=1/3) + dbinom(x=3,size=3, prob=1/3)
```

### Question 2:
suppose that $F$ denotes that a randomly selected individual is female and $B$ denotes that a randomly selected
individual has blonde hair. It is known that $P(F) = 0.57, P(B)=0.21$ and $P(F \cup B)=0.60$. Find the conditional
probability that an individual is female given that we know they have blonde hair. Give your answer to two decimal
places.

```{r}
F <- 0.57
M <- 0.43
B <- 0.21
NB <- 0.79
ForB <- 0.6  # not entirely sure of the relevance of this

FandB <- F + B - ForB  # P(female) plus P(blonde), less the intersection
# FandB  # female and blonde
FandB / B  # the probability that the person is female given they are blonde
```

### Question 3:
Let $X$ be a continuous random variable with density

$$f_X (x) = \dfrac{x}{2}$$

for $0 < x < c$. Find the value of the constant $c$ that makes $f_X(x)$ a valid probability density function. Enter the
exact value.

*this looks like it relates to the probability density function theorem example*
two conditions

```{r}
# Plot the density function
pdf <- function(x) x/2
# Show that the density function integrates to one
integrate(pdf,0,2)
```

### Question 4: (true/false)
Suppose there are two events $A$ and $B$ such that $P(A^c)=0.2$ and $P(B)=0.3$. The two events $A$ and $B$ can be disjoint.

```{r}

```

### Question 5: (multi choice)
Let $X$ has a probability density function

$$f(x)= \theta e^{-\theta x}$$

for $0 < x < \infty$ and $\theta > 0$ a positive constant. Find the probability density function of

$$Y = \sqrt{X}$$

The domain must also be specified in you answer.

*pdf*

```{r}

```

### Question 6: (multichoice)
Let the continuous random vector $(X,Y)$ have joint density

$$f(x,y)=e^{-x}, 0<y<x<\infty$$

Find the conditional density of $X$ conditioned on $Y=y$

```{r}

```

### Question 7:
For the following density function of $X$

$$f(x) = \begin{Bmatrix}
2(x-1) & \text{if} & 1 < x < 2, \\
0 & & \text{zero elsewhere}
\end{Bmatrix}$$

*this looks like the geometric distribution probability mass function*

compute $P(X>1.3)$. your answer should be to two decimal places.
```{r}

```

### Question 8:
Let the number of insurance claims in a particular hour for an insurance company have a poisson distribution with probability mass function
$$f(x, \lambda) = \dfrac{e^{-\lambda}\lambda^x}{x!}, for x \in 0,1,2,..., \lambda >0$$
You are asked to **find the smallest value of the mean (which equals $\lambda$) of this distribution that ensures the probability that a randomly chosen hour has zero insurance claims to be smaller than 0.20**. Your answer should be to two decimal places.
*poisson pmf*

# (this isn't correct)

```{r}
ppois(0, 1.61)
```

### Question 9:
Let $T$ have probability density function
$$f_T (t) = \dfrac{nt^{nt^{n-1}}}{\theta^n}$$
for $0<t< \theta$ where $\theta > 0$ and $n$ is a known integer. Compute the mean of T.
*mean is the square root of variance*
# problem
```{r}
x <- n*t^(n-1)/a^x
moment(x, order=1, center=FALSE, absolute=FALSE, na.rm=FALSE)
```

### Question 10: (multichoice)

Let $T$ have probability density function

$$f_T (t) = \dfrac{nt^{n-1}}{\theta^n}$$

for $0 < t < \theta$ where $\theta > 0$ and $n$ is a known integer. Compute the variance of $T$

**Hint**: It is known that for any random variable $X$ we have $\text{Var}(X) = E(X^2) - (E(X))^2$


```{r}

```
